---
title: "Intro to ggplot2"
author: "Travis Blimkie"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    toc: yes
    code_folding: "show"
---

***

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, comment = "")
```

<br>

> These materials have been taken from the Software Carpentry: R Novice Lesson.
You can find the original materials
[here](https://github.com/swcarpentry/r-novice-gapminder).

<br>

Plotting our data is one of the best ways to quickly explore it and the various
relationships between variables.

There are three main plotting systems in R, the [base plotting
system](http://www.statmethods.net/graphs/index.html), the
[lattice](http://www.statmethods.net/advgraphs/trellis.html) package, and the
[ggplot2](http://www.statmethods.net/advgraphs/ggplot2.html) package.


Today we'll be learning about the ggplot2 package, because it is the most
effective for creating publication quality graphics.
ggplot2 is built on the grammar of graphics, the idea that any plot can be
expressed from the same set of components: a **data** set, a **coordinate
system**, and a set of **geoms**, the visual representation of data points.

The key to understanding ggplot2 is thinking about a figure in layers. This idea
may be familiar to you if you have used image editing programs like Photoshop,
Illustrator, or Inkscape.

<br>

First, let's make sure everything's loaded:
```{r}
# Download the packages
# install.packages(c("tidyverse", "gapminder"))

# Load the packages for use
library(tidyverse)
library(gapminder)
```

<br>

## Simple ggplot
Let's start off with an example:

```{r}
ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp)) +
  geom_point()
```

<br>

So the first thing we do is call the `ggplot` function. This function lets R
know that we're creating a new plot, and any of the arguments we give the
`ggplot` function are the *global* options for the plot: they apply to **all**
layers on the plot.

We've passed in two arguments to `ggplot`. First, we tell `ggplot` what data we
want to show on our figure, in this example the gapminder data we read in
earlier. For the second argument we passed in the `aes` function, which
tells `ggplot` how variables in the **data** map to *aesthetic* properties of
the figure, in this case the **x** and **y** locations. Here we told `ggplot` we
want to plot the "gdpPercap" column of the gapminder data frame on the x-axis, and
the "lifeExp" column on the y-axis.

By itself, the call to `ggplot` isn't enough to draw a figure:

```{r}
ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp))
```

<br>

We need to tell `ggplot` how we want to visually represent the data, which we
do by adding a new **geom** layer. In our example, we used `geom_point`, which
tells `ggplot` we want to visually represent the relationship between **x** and
**y** as a scatterplot of points:

```{r}
ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp)) +
  geom_point()
```

<br>

### Challenge 1
Modify the example so that the figure shows how life expectancy has
changed over time:
```{r, eval=FALSE}
# Previous example:
ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp)) + 
  geom_point()
```
Hint: the gapminder dataset has a column called "year", which should appear
on the x-axis.

### Solution to Challenge 1
Here is one possible solution:
```{r}
ggplot(data = gapminder, aes(x = year, y = lifeExp)) + 
  geom_point()
```

<br>

### Challenge 2
In the previous examples and challenge we've used the `aes` function to tell
the scatterplot **geom** about the **x** and **y** locations of each point.
Another *aesthetic* property we can modify is the point *colour*. Modify the
code from the previous challenge to **colour** the points by the "continent"
column. What trends do you see in the data? Are they what you expected?

### Solution to Challenge 2
```{r}
ggplot(data = gapminder, aes(x = year, y = lifeExp, colour = continent)) +
 geom_point()
```

<br>

## Layers
Using a scatterplot probably isn't the best for visualizing change over time.
Instead, let's tell `ggplot` to visualize the data as a line plot. Here we'll
also shorten our code by omitting some of the declarations (e.g. x = year). By
default, the first object passed to `ggplot()` is assumed to be the `data`
argument. Similarly, the first two column names of `aes()` are assumed to be the
x and y arguments.

```{r}
ggplot(gapminder, aes(year, lifeExp, by = country, colour = continent)) +
  geom_line()
```

<br>

Instead of adding a `geom_point` layer, we've added a `geom_line` layer. We've
added the **by** *aesthetic*, which tells `ggplot` to draw a line for each
country.

But what if we want to visualize both lines and points on the plot? We can
simply add another layer to the plot:

```{r}
ggplot(gapminder, aes(year, lifeExp, by = country, colour = continent)) +
  geom_line() + 
  geom_point()
```

<br>

It's important to note that each layer is drawn on top of the previous layer. In
this example, the points have been drawn *on top of* the lines. Here's a
demonstration:

```{r}
ggplot(gapminder, aes(year, lifeExp, by = country)) +
  geom_line(aes(colour = continent)) + 
  geom_point()
```

<br>

In this example, the *aesthetic* mapping of **colour** has been moved from the
global plot options in `ggplot` to the `geom_line` layer so it no longer applies
to the points. Now we can clearly see that the points are drawn on top of the
lines.

### Tip: Setting an aesthetic to a value instead of a mapping
So far, we've seen how to use an aesthetic (such as **colour**) as a *mapping* to
a variable in the data. For example, when we use `geom_line(aes(colour =
continent))`, ggplot will give a different colour to each continent. But what if
we want to change the colour of all lines to blue? You may think that
`geom_line(aes(colour = "blue"))` should work, but it doesn't. Since we don't
want to create a mapping to a specific variable, we simply move the colour
specification outside of the `aes()` function, like this: `geom_line(colour =
"blue")`.

<br>

### Challenge 3
Switch the order of the point and line layers from the previous example. What
happened?

### Solution to Challenge 3
Switch the order of the point and line layers from the previous example. What
happened?
```{r}
ggplot(gapminder, aes(year, lifeExp, by = country)) +
  geom_point() + 
  geom_line(aes(colour = continent))
```

<br>

The lines now get drawn over the points!

<br>

## Transformations and statistics
ggplot2 also makes it easy to overlay statistical models over the data. To
demonstrate we'll go back to our first example:

```{r}
ggplot(gapminder, aes(gdpPercap, lifeExp, colour = continent)) +
  geom_point()
```

<br>

Currently it's hard to see the relationship between the points due to some strong
outliers in GDP per capita. We can change the scale of units on the x axis using
the *scale* functions. These control the mapping between the data values and
visual values of an aesthetic. We can also modify the transparency of the
points, using the *alpha* function, which is especially helpful when you have
a large amount of data which is very clustered.

```{r}
ggplot(gapminder, aes(gdpPercap, lifeExp)) +
  geom_point(alpha = 0.5) + 
  scale_x_log10()
```

<br>

The `log10` function applied a transformation to the values of the gdpPercap
column before rendering them on the plot, so that each multiple of 10 now only
corresponds to an increase in 1 on the transformed scale, e.g. a GDP per capita
of 1,000 is now 3 on the y axis, a value of 10,000 corresponds to 4 on the y
axis and so on. This makes it easier to visualize the spread of data on the
x-axis.

### Reminder: Setting an aesthetic to a value instead of a mapping
Notice that we used `geom_point(alpha = 0.5)`. As the previous tip mentioned, using a setting outside of the `aes()` function will cause this value to be used for all points, which is what we want in this case. But just like any other aesthetic setting, *alpha* can also be mapped to a variable in the data. For example, we can give a different transparency to each continent with `geom_point(aes(alpha = continent))`.

<br>

We can fit a simple relationship to the data by adding another layer,
`geom_smooth`:

```{r}
ggplot(gapminder, aes(gdpPercap, lifeExp)) +
  geom_point() + 
  scale_x_log10() + 
  geom_smooth(method = "lm")
```

<br>

We can make the line thicker by *setting* the **size** aesthetic in the
`geom_smooth` layer:

```{r}
ggplot(gapminder, aes(gdpPercap, lifeExp)) +
  geom_point() + 
  scale_x_log10() + 
  geom_smooth(method = "lm", size = 1.5)
```

<br>

There are two ways an *aesthetic* can be specified. Here we *set* the **size**
aesthetic by passing it as an argument to `geom_smooth`. Previously in the
lesson we've used the `aes` function to define a *mapping* between data
variables and their visual representation.

### Challenge 4a
Modify the colour and size of the points on the point layer in the previous
example. Hint: do not use the `aes` function.

### Solution to challenge 4a
```{r}
ggplot(gapminder, aes(gdpPercap, lifeExp)) +
  geom_point(size = 3, colour = "orange") + 
  scale_x_log10() +
  geom_smooth(method = "lm", size = 1.5)
```

<br>

### Challenge 4b
Modify your solution to Challenge 4a so that the
points are now a different shape and are coloured by continent with new
trendlines.  Hint: The colour argument can be used inside the aesthetic.

### Solution to Challenge 4b
Modify Challenge 4 so that the points are now a different shape and are
coloured by continent with new trendlines.
Hint: The colour argument can be used inside the aesthetic.
```{r}
ggplot(gapminder, aes(gdpPercap, lifeExp, colour = continent)) +
  geom_point(size = 2, alpha = 0.5) + 
  scale_x_log10() +
  geom_smooth(method = "lm", size = 1.5)
```

<br>

## Multi-panel figures
Earlier we visualized the change in life expectancy over time across all
countries in one plot. Alternatively, we can split this out over multiple panels
by adding a layer of **facet** panels. Here we will first make a smaller version
of the data to make it easier to work with. The `%in%` operator just means that
country must be equal to one of the four specified countries.

```{r}
gapminder_small <- gapminder %>% 
  filter(country %in% c("Canada", "United States", "France", "Australia"))

ggplot(gapminder_small, aes(year, lifeExp, colour = continent)) +
  geom_line() + 
  facet_wrap(~country)
```

<br>

The `facet_wrap` layer took a "formula" as its argument, denoted by the tilde
(`~`). This tells R to draw a panel for each unique value in the country column
of the `gapminder_small` data.

<br>

## Modifying text
To clean this figure up for a publication we need to change some of the text
elements. The x-axis is too cluttered, and the y axis should read "Life
expectancy", rather than the column name in the data frame.

We can do this by adding a couple of different layers. The **theme** layer
controls the axis text, and overall text size. Labels for the axes, plot title
and any legend can be set using the `labs` function. Legend titles are set using
the same names we used in the `aes` specification. Thus below the colour legend
title is set using `colour = "Continent"`, while the title of a fill legend
would be set using `fill = "MyTitle"`.

```{r}
gapminder_small <- gapminder %>% 
  filter(country %in% c("Canada", "United States", "France", "Australia"))

ggplot(gapminder_small, aes(year, lifeExp, colour = continent)) +
  geom_line() + 
  facet_wrap(~country) +
  labs(x = "Year", y = "Life expectancy", title = "Figure 1", colour = "Continent") +
  theme_bw() +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
```

<br>

## Exporting the plot
The `ggsave()` function allows you to export a plot created with ggplot. You can
specify the dimension and resolution of your plot by adjusting the appropriate
arguments (`width`, `height` and `dpi`) to create high quality graphics for
publication. In order to save the plot from above, we first assign it to a
variable `lifeExp_plot`, then tell `ggsave` to save that plot in `png` format to
a directory called `results`. (Make sure you have a `results/` folder in your
working directory.)

```{r}
lifeExp_plot <- ggplot(gapminder_small, aes(year, lifeExp, colour = continent)) +
  geom_line() + 
  facet_wrap(~country) +
  labs(x = "Year", y = "Life expectancy", title = "Figure 1", colour = "Continent") +
  theme_bw() +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())

ggsave(plot = lifeExp_plot, filename = "lifeExp.png", units = "cm", width = 12, height = 10, dpi = 300)
```

<br>

There are two nice things about `ggsave`. First, it defaults to the last plot,
so if you omit the `plot` argument it will automatically save the last plot you
created with `ggplot`. Secondly, it tries to determine the format you want to
save your plot in from the file extension you provide for the filename (for
example `.png` or `.pdf`). If you need to, you can specify the format explicitly
in the `device` argument.

<br>

## Summary
This is a taste of what you can do with ggplot2. RStudio provides a really
useful [cheat sheet](http://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf)
of the different layers available, and more extensive documentation is available
on the [ggplot2 website](http://docs.ggplot2.org/current/). Finally, if you have
no idea how to change something, a quick Google search will usually send you to
a relevant question and answer on Stack Overflow with reusable code to modify!

<br>

### Challenge 5
Create a density plot of GDP per capita, filled by continent.

Advanced:

* Transform the x axis to better visualise the data spread
* Add a facet layer to panel the density plots by year
* Make the y-axis range independent for each facet (year)

### Solution to Challenge 5
```{r}
ggplot(gapminder, aes(gdpPercap, fill = continent)) +
  geom_density(alpha = 0.6) + 
  facet_wrap(~year, scales = "free_y") + 
  scale_x_log10()
```

<br>
