---
title: "Importing Data, Basic Data Manipulation/Viewing"
output: html_notebook
---

## Obtaining Data

First, let's download the data.

```{r}
ROTK <- read.csv(file="https://raw.githubusercontent.com/UBC-R-Study-group/introRworkshop_Sept19/master/data/The_Return_Of_The_King.csv")
```

We could also load directly from a file.

```{r}
FOTR <- read.csv(file="/Users/arnavsood/sandbox/introRworkshop_Sept19/data/The_Fellowship_Of_The_Ring.csv")
```

And we can load directly from an Excel file.

```{r}
library("readxl")
TTT <- read_excel("/Users/arnavsood/sandbox/introRworkshop_Sept19/data/The_Two_Towers.xlsx")
```

## Viewing Data

The basic way to see a small data table is with the ``print`` function 

```{r}
print(ROTK)
```

You can also use the ``View`` function to open an RStudio window. 

```{r}
View(ROTK)
```

## Editing Data 

By default, data tables aren't open to modification. This helps preserve the integrity of the data. 

If you want to edit, you can use the ``edit()`` function. 

```{r}
TTT <- edit(TTT)
```

## Initial Inspection of Data 

One nice function we can use is ``str``.

```{r}
str(FOTR)
```

We can also inspect individual columns. 

Notice that RStudio helps us along by autofilling the various columns.

```{r}
str(FOTR$Race)
```

## Quantitatively Analyzing Data

The next step is figuring out how to analyze data. 

A simple thing to do is geneate summary statistics. 

```{r}
summary(ROTK)
```

```{r}
summary(TTT)
```

We could try selecting individual columns.

```{r}
ROTK$Race
```

We can also subset the data quantitatively, by index.

```{r}
ROTK[2, 2]
```

```{r}
ROTK[10]
```

Why did that happen?


